Grammar-Based Concept Alignment for Domain-Specific Machine Translation

### 2021-05-16 paper/paper.tex
last few major changes; only final review left

% input
Here, the input consists of a list of priority-sorted \textit{alignment criteria}, i.e. rules to determine whether two dependency trees should be aligned with each other, and a pair $(t,u)$ of UD trees to align.
An example alignment criterion is sameness of syntactic label, for instance, that subjects are aligned with subjects and objects with objects; the details will be discussed in Section \ref{criteria},  
From an implementation point of view, UD trees are, rose trees (trees with arbitrary numbers of branches) where each node represents a word with its dependents as subtrees.
The rose tree is is easily obtained from the CoNLL-U notation that UD parsers produce.

\begin{figure*}[h]
  \centering
  \begin{subfigure}{.23\textwidth}
      \centering
      \scriptsize
      \setlength{\unitlength}{0.16mm}
      \begin{picture}(371.0,130.0)
          \put(0.0,0.0){this}
          \put(46.0,0.0){is}
          \put(83.0,0.0){a}
          \put(120.0,0.0){dependency}
          \put(220.0,0.0){tree}
          \put(0.0,15.0){{\tiny PRON}}
          \put(46.0,15.0){{\tiny AUX}}
          \put(83.0,15.0){{\tiny DET}}
          \put(120.0,15.0){{\tiny NOUN}}
          \put(220.0,15.0){{\tiny NOUN}}
          \put(120.0,30.0){\oval(218.63636363636363,133.33333333333334)[t]}
          \put(10.681818181818187,35.0){\vector(0,-1){5.0}}
          \put(105.0,99.66666666666667){{\tiny nsubj}}
          \put(143.0,30.0){\oval(172.27586206896552,100.0)[t]}
          \put(56.86206896551724,35.0){\vector(0,-1){5.0}}
          \put(128.0,83.0){{\tiny cop}}
          \put(161.5,30.0){\oval(134.8102189781022,66.66666666666667)[t]}
          \put(94.0948905109489,35.0){\vector(0,-1){5.0}}
          \put(146.5,66.33333333333334){{\tiny det}}
          \put(180.0,30.0){\oval(97.0,33.333333333333336)[t]}
          \put(131.5,35.0){\vector(0,-1){5.0}}
          \put(165.0,49.66666666666667){{\tiny compound}}
          \put(235.0,130.0){\vector(0,-1){100.0}}
          \put(240.0,120.0){{\tiny root}}
      \end{picture}
  \end{subfigure}
  \begin{subfigure}{.38\textwidth}
    \centering
    \scriptsize
      \begin{verbatim}
      1	this	this	PRON	_	_	5	nsubj	_	_
      2	is	be	AUX	_	_	5	cop	_	_
      3	a	a	DET	_	_	5	det	_	_
      4	dependency	dependency	NOUN	_	_	5	
        compound	_	_
      5	tree	tree	NOUN	_	_	0	root	_	_
      \end{verbatim}
  \end{subfigure}%
  \begin{subfigure}{.38\textwidth}
    \centering
    \scriptsize
      \begin{verbatim}
5	tree	tree	NOUN	_	_	0	root	_	_
  4	dependency	dependency	NOUN	_	_	5	
    compound	_	_
  2	is	be	AUX	_	_	5	cop	_	_
  3	a	a	DET	_	_	5	det	_	_
  1	this	this	PRON	_	_	5	nsubj	_	_
      \end{verbatim}
  \end{subfigure}%
  \caption[The graphical representation of a UD tree alongside its CoNNL-U and sorted rose tree representation. Optional CoNNL-U fields are left blank.]{The graphical representation of a UD tree alongside its CoNNL-U and rose tree representation. Optional CoNNL-U fields are left blank.} 
\end{figure*}

% algorithm description
As a first step, the program checks whether the two full sentence trees can be aligned with each other, i.e. if they match the alignment criteria. 
If this is the case, they are added to a collection of alignments, which are represented as pairs of UD (sub)trees associated with some metadata, such as the id of the sentence they were extracted from.
Such a collection is what the function will return after aligning all the dependency subtrees.
The same procedure is applied recursively to all pairs of immediate subtrees of each sentence, until the leaves are reached or alignment is no longer possible due to lack of matching criteria.
Subtrees are sorted based on their dependency label, to give higher priority to pairs whose heads have the same label.

@@

% details
Of course, \texttt{fast\_align} does not make use of the information present in the CoNLL-U files except with regards to tokenization. 
On the other hand, the relatively large size of the PUD treebanks makes it possible also to train the statistical tool on the full dataset instead of just using the chosen 100-sentences subset, allowing for a fairer comparison. 
In both cases, \texttt{fast\_align} is run with the recommended parameters and the CE program is configured only to extract one-to-many and many-to-one word alignments, given that \texttt{fast\_align} does not align larger phrases.

>>>>

% input
Here, the input consists of a list of priority-sorted \textit{alignment criteria}, i.e. rules to determine whether two dependency trees should be aligned with each other, and a pair $(t,u)$ of UD trees to align.
An example alignment criterion is sameness of syntactic label, for instance, that subjects are aligned with subjects and objects with objects; the details will be discussed in Section \ref{criteria}.  
From an implementation point of view, UD trees are rose trees (trees with arbitrary numbers of branches) where each node represents a word with its dependents as subtrees.
The rose tree is is easily obtained from the CoNLL-U notation that UD parsers produce.

\begin{figure*}[h]
  \centering
  \scriptsize
  \begin{subfigure}{.30  \textwidth}
    \setlength{\unitlength}{0.14mm}
\begin{picture}(257.0,70.0)
  \put(0.0,0.0){She}
  \put(46.0,0.0){studies}
  \put(119.0,0.0){consistently}
  \put(0.0,15.0){{\tiny PRON}}
  \put(46.0,15.0){{\tiny VERB}}
  \put(119.0,15.0){{\tiny ADV}}
  \put(33.0,30.0){\oval(39.47826086956522,33.333333333333336)[t]}
  \put(13.26086956521739,35.0){\vector(0,-1){5.0}}
  \put(18.0,49.66666666666667){{\tiny nsubj}}
  \put(102.5,30.0){\oval(68.89041095890411,33.333333333333336)[t]}
  \put(136.94520547945206,35.0){\vector(0,-1){5.0}}
  \put(87.5,49.66666666666667){{\tiny advmod}}
  \put(61.0,70.0){\vector(0,-1){40.0}}
  \put(66.0,60.0){{\tiny root}}
\end{picture}
\\ \\
\begin{picture}(259.0,90.0)
  \put(0.0,0.0){Lei}
  \put(46.0,0.0){studia}
  \put(110.0,0.0){con}
  \put(147.0,0.0){costanza}
  \put(0.0,15.0){{\tiny PRON}}
  \put(46.0,15.0){{\tiny NOUN}}
  \put(110.0,15.0){{\tiny ADP}}
  \put(147.0,15.0){{\tiny NOUN}}
  \put(33.0,30.0){\oval(39.47826086956522,33.333333333333336)[t]}
  \put(13.26086956521739,35.0){\vector(0,-1){5.0}}
  \put(18.0,49.66666666666667){{\tiny nsubj}}
  \put(138.5,30.0){\oval(28.89189189189189,33.333333333333336)[t]}
  \put(124.05405405405405,35.0){\vector(0,-1){5.0}}
  \put(123.5,49.66666666666667){{\tiny case}}
  \put(116.5,30.0){\oval(98.02970297029702,66.66666666666667)[t]}
  \put(165.5148514851485,35.0){\vector(0,-1){5.0}}
  \put(101.5,66.33333333333334){{\tiny nmod}}
  \put(61.0,90.0){\vector(0,-1){60.0}}
  \put(66.0,80.0){{\tiny root}}
\end{picture}
  \end{subfigure}
  \begin{subfigure}{.30  \textwidth}
    \begin{verbatim}

1 She PRON 2 nsubj
2 studies VERB 0 root    
3 consistently ADV 2 advmod

1 Lei PRON 2 nsubj
2 studia VERB 0 root
3 con ADP 4 case
4 costanza NOUN 2 obl
    \end{verbatim}
  \end{subfigure}
  \begin{subfigure}{.30 \textwidth}
    \begin{verbatim}

2 studies VERB 0 root  
  3 consistently ADV 2 advmod
  1 She PRON 2 nsubj

2 studia VERB 0 root
  1 Lei PRON 2 nsubj
  4 costanza NOUN 2 obl
    3 con ADP 4 case
    \end{verbatim}
  \end{subfigure}
  \caption{The graphical and simplified CoNNL-U representation of a pair of UD sentences. Below, the corresponding sorted rose trees. With the default parameters, the resulting alignments are: $\langle$\textit{She studies consistently, Lei studia con costanza}$\rangle$, $\langle$\textit{studies, studia}$\rangle$, $\langle$\textit{she, lei}$\rangle$ and $\langle$\textit{consistently, con costanza}.$\rangle$} 
\end{figure*}

% algorithm description
As a first step, the program checks whether the two full sentence trees can be aligned with each other, i.e. if they match the alignment criteria. 
In the case of the example criterion discussed above, this means that their roots are labelled the same.
If this is the case, they are added to a collection of alignments, which are represented as pairs of UD (sub)trees associated with some metadata, such as the id of the sentence they were extracted from.
Such a collection is what the function will return after aligning all the dependency subtrees.
The same procedure is applied recursively to all pairs of immediate subtrees of each sentence, until the leaves are reached or alignment is no longer possible due to lack of matching criteria.
Subtrees are sorted based on their dependency label to give higher priority to pairs whose heads have the same label (cf. \textsc{sort} in the pseudocode).

@@

% details
Of course, \texttt{fast\_align} does not make use of the information present in the CoNLL-U files except with regards to tokenization. 
On the other hand, the relatively large size of the PUD treebanks makes it possible also to train the statistical tool on the full dataset instead of just using the chosen 100-sentences subset, allowing for a fairer comparison. 
In both cases, \texttt{fast\_align} is run with the recommended parameters and the CE program is configured only to extract one-to-many and many-to-one word alignments, given that \texttt{fast\_align} does not align larger phrases. 
This explains CE's seemingly low recall.
To get a better idea, Table \ref{pud_fast} can be compared with Table \ref{raw_fast}, which summarizes the results of an experiment where no constraints are placed on the size of the extracted alignments.
